---
title: "Coursera Statistical Inference - Course Project II"
author: "Matt Flor"
date: "September 20, 2015"
output: 
  pdf_document:
    fig_caption: yes
    fig_height: 3
    fig_width: 5
    number_sections: yes
subtitle: 'Tooth Growth: A Basic Exploratory Data Analysis'
---

**Note that this document is fully reproducible, so that I am not giving a 3 pages + 3 pages supplementary report but rather a 6 page report with code and text together.**

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

First, we load required `R` packages and the data.
The `ToothGrowth` dataset describes the effect of vitamin C at three **dose** levels on tooth growth (variable **len**) in guinea pigs with each of two delivery methods (orange juice or pure Vitamin C (ascorbic acid)).
We convert the dose variable to a factor, and rename the levels of the **suppl** variable (i.e. the delivery method) to make them easier to understand.

```{r data}
library(plyr)
library(dplyr)
library(ggplot2)
library(pander)
library(datasets)
data("ToothGrowth")
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
ToothGrowth$supp <- plyr::revalue(ToothGrowth$supp, 
                                  c("VC" = "Vitamin C", "OJ" = "Orange juice"))
set.caption("First rows of the dataset.")
pander(head(ToothGrowth))
```

Now we can look at some simple summary statistics for the three variable, tooth length (len), delivery methods (supp), and Vitamin C dose (dose).

```{r summaries}
set.caption("Summary of tooth lengths.")
pander(summary(ToothGrowth$len))
set.caption("Group sizes for the two delivery methods.")
pander(summary(ToothGrowth$supp))
set.caption("Group sizes for the dose levels of Vitamin C (in mg).")
pander(summary(ToothGrowth$dose))
```

```{r comparisons}
comps <- ToothGrowth %>% 
    group_by(supp, dose) %>% 
    summarise(mean = mean(len), sd = sd(len))
set.caption("Mean tooth length and standard deviation for each of the 
            combination of delivery method and dose.")
pander(comps)
```


