---
title: "Severe Weather Events: Harm to Population Health and Economic Consequences"
author: "Matt Flor"
date: "August 21, 2015"
output:
    html_document:
        number_sections: yes
        fig_caption: yes
        toc: yes
---

# Synopsis  

§1 Severe weather events may harm the population health and may damage property and crops.  
§2 In this analysis, the National Oceanic and Atmospheric Administration's Storm Data is used to assess these health and economic consequences.   
§3 In summary, tornados, excessive heat, flood, thunderstorm wind, and lightning pose the greatest risk to population health. Tornados injure the most people by far, whereas excessive heat claims most fatalities.  
§4 In general, injury numbers are one order of magnitude larger than fatality numbers.  
§5 There is considerable variability across years, in particular for tornado, flood, and excessive heat events.  
§6 Economic consequences are mainly due to property damage, with flood, hurricane, storm surge, tornado, and hail events contributing the most over the years.  
§7 A notable exception are drought events which damage crops the most.  
§8 Variability across the years is mostly produced by flood and hurricane events.  


# Data

*Storm Data* is an official publication of the National Oceanic and Atmospheric Administration (NOAA).
It documents

> a. The occurrence of storms and other significant weather phenomena having sufficient intensity to cause loss of life, injuries, significant property damage, and/or disruption to commerce;  
> b. Rare, unusual, weather phenomena that generate media attention, such as snow flurries in South Florida or the San Diego coastal area; and  
> c. Other significant meteorological events, such as record maximum or minimum temperatures or precipitation that occur in connection with another event.  
> *From the disclaimer in the [Storm Data Documentation][StormDataDoc] by the National Weather Service*

The [Storm Data Documentation][StormDataDoc] contains detailed information on the data collection process.

In addition, a [Storm Events FAQ][StormDataFAQ] by the National Climatic Data Center is available.

[StormDataDoc]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
[StormDataFAQ]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf

For this analysis, the storm database can be downloaded from the Coursera web site.

```{r download.data, echo=TRUE}
fileName <- "StormData.csv.bz2"
if (!file.exists(fileName)) {
    dataURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(dataURL, destfile = fileName, method = "curl")
}
```


# Data Processing

In this section, the data processing steps are detailed (in word and actual code).

Note that if you are not interested in reproducing this report you may skip this section and go straight to the [Results](#results) part.

## Software and Notes on Reproducibility

This report was generated on a PC running Ubuntu Linux 14.04 and the statistical software R (version 3.2.2) and is fully reproducible.

In order to reproduce the report, the following R packages must be installed:

```{r load.libraries, echo=TRUE, message=FALSE, warning=FALSE}
library(plyr)
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)
library(ggplot2)
library(gridExtra)
library(knitr)
```

Detailed information on the versions used can be taken from the following session info that was recorded while generating this report.

```{r session.info, echo=TRUE}
sessionInfo()
```

To generate the HTML report, simply run the command `knitr::knit("Severe_Weather_Events.Rmd")`.
Make sure that your global R environment is empty if you generate the report for the first time.


## Loading

For the analysis, the database is loaded into a data frame `storms.df`.
Variables that are not necessary for the analysis are dropped, and a new `year` variable is added.

```{r read.data, echo=TRUE, cache=TRUE}
if (!exists("storms.df")) storms.df <- tbl_df(read.csv("StormData.csv.bz2"))
storms.df <- storms.df %>% 
    mutate(year = year(mdy_hms(BGN_DATE))) %>% 
    select(REFNUM, year, EVTYPE, FATALITIES, INJURIES,
           PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
```

## Excluding Non-Relevant Records

### Years

For this analysis, in order to reliably summarize and compare data, only data from 1996 onwards is included.
According to information on the National Climatic Data Center' website, records from years before 1996 either do not include the full set of event types used today ([1][NCDCevtype]) or have been extracted from unformatted text files and "contain many inconsistencies in the spelling of event types" ([2][NCDCsuppl]).

[NCDCevtype]: http://www.ncdc.noaa.gov/stormevents/details.jsp?type=eventtype
[NCDCsuppl]: http://www.ncdc.noaa.gov/stormevents/details.jsp?type=supplemental

```{r filter.years, echo=TRUE}
storms.df <- storms.df %>% 
    filter(year > 1995)
```


### Events without Actual Harm to Population Health or Damage

We exclude records of events that neither harmed population health nor damaged properties or crops.

```{r filter.harm.dmg, echo=TRUE}
storms.df <- storms.df %>% 
    filter(FATALITIES != 0 | INJURIES != 0 | PROPDMG != 0 | CROPDMG != 0)
```

### Summary Records

We exclude rows that provide summaries for certain days or event types to avoid double counting, and drop unused levels.

```{r filter.summaries, echo=TRUE}
storms.df <- storms.df %>% 
    filter(!grepl("[sS]ummary", EVTYPE))
storms.df <- droplevels(storms.df)
```

## Cleaning Up Event Types

The dataset contains many records with event types that are not permitted according to section 2.1 of the [Storm Data Documentation][StormDataDoc].

We do our best to match non-permitted event types to permitted ones.

```{r permitted.evtypes, echo=TRUE}
permitted.evtypes <- toupper(c(
    "ASTRONOMICAL LOW TIDE", "AVALANCHE", 
    "BLIZZARD", 
    "COASTAL FLOOD", "COLD/WIND CHILL", 
    "DEBRIS FLOW", "DENSE FOG", "DENSE SMOKE", "DROUGHT", "DUST DEVIL", "DUST STORM", 
    "EXCESSIVE HEAT", "EXTREME COLD/WIND CHILL", 
    "FLASH FLOOD", "FLOOD", "FROST/FREEZE", "FUNNEL CLOUD", "FREEZING FOG", 
    "HAIL", "HEAT", "HEAVY RAIN", "HEAVY SNOW", "HIGH SURF", "HIGH WIND", "HURRICANE (TYPHOON)", 
    "ICE STORM", 
    "LAKE-EFFECT SNOW", "LAKESHORE FLOOD", "LIGHTNING", 
    "MARINE HAIL", "MARINE HIGH WIND", "MARINE STRONG WIND", "MARINE THUNDERSTORM WIND", 
    "RIP CURRENT", 
    "SEICHE", "SLEET", "STORM SURGE/TIDE", "STRONG WIND", 
    "THUNDERSTORM WIND", "TORNADO", "TROPICAL DEPRESSION", "TROPICAL STORM", "TSUNAMI", 
    "VOLCANIC ASH", 
    "WATERSPOUT", "WILDFIRE", "WINTER STORM", "WINTER WEATHER"
))
```

A range of regular expression matching and replacing is performed.
Any remaining non-permitted event type is assigned a new "UNMATCHED" type so that we can later judge potential problems.

```{r matching.evtypes, echo=TRUE, results="asis"}
mgsub <- function(pattern, replacement, x, ...) {
    # hat tip for this function goes to Theodore Lytras, http://stackoverflow.com/a/15254254
    if (length(pattern) != length(replacement)) {
        stop("pattern and replacement do not have the same length.")
    }
    result <- x
    for (i in 1:length(pattern)) {
        result <- gsub(pattern[i], replacement[i], result, ...)
    }
    result
}

patterns <- c(" {0,}\\(?G?\\d\\d\\)?",
              "TSTM",
              "^HURRICANE(/TYPHOON| EDOUARD|$)",
              "^TYPHOON$",
              "^(URBAN/SML STREAM FLD|RIVER FLOOD(ING)?)$",
              "/HAIL",
              "^(LAND|MUD|ROCK )SLIDE$",
              "^STORM SURGE$",
              "^WILD/FOREST FIRE$",
              "^RIP CURRENTS$",
              "^EXTREME (COLD|WINDCHILL)$",
              "^(LIGHT )?SNOW$|^WINTER WEATHER[ /]MIX$|WINTRY MIX|IC(E|Y) (ON )?ROADS|FREEZING DRIZZLE$",
              "^(AGRICULTURAL |DAMAGING )?FREEZE$|(LIGHT )?FREEZING RAIN$|^GLAZE$",
              "^(TIDAL|COASTAL) FLOODING(/EROSION)?$",
              "^(COLD)$",
              "^FOG$",
              "^(WIND$|(STRONG|GUSTY) WINDS?|(DRY|WET) MICROBURST)$",
              "^EXCESSIVE SNOW$",
              "^(HEAVY SURF(/HIGH SURF)?|(HIGH|ROUGH) SEAS)$",
              "^SMALL HAIL$",
              "^MIXED PRECIP(ITATION)?$|^RAIN(/SNOW)?$")
replacements <- c("",
                  "THUNDERSTORM",
                  "HURRICANE (TYPHOON)",
                  "HURRICANE (TYPHOON)",
                  "FLOOD",
                  "",
                  "DEBRIS FLOW",
                  "STORM SURGE/TIDE",
                  "WILDFIRE",
                  "RIP CURRENT",
                  "EXTREME COLD/WIND CHILL",
                  "WINTER WEATHER",
                  "FROST/FREEZE",
                  "COASTAL FLOOD",
                  "COLD/WIND CHILL",
                  "DENSE FOG",
                  "STRONG WIND",
                  "HEAVY SNOW",
                  "HIGH SURF",
                  "HAIL",
                  "HEAVY RAIN")

storms.df <- storms.df %>% 
    mutate_each(funs(toupper(as.character(.))), EVTYPE) %>% 
    mutate_each(funs(str_trim), EVTYPE)
storms.df$EVTYPE <- mgsub(patterns, replacements, storms.df$EVTYPE)
storms.df$EVTYPE[!(storms.df$EVTYPE %in% permitted.evtypes)] <- "UNMATCHED"
storms.df$EVTYPE <- as.factor(storms.df$EVTYPE)
```

## Damage Calculation

Damage information is stored in four different variables:

- `PROPDMG`: Property damage number
- `PROPDMGEXP`: Letter signifying the magnitude of the property damage number
- `CROPDMG`: Crop damage number
- `CROPDMGEXP`: Letter signifying the magnitude of the crop damage number

The letters are  "K" for thousands, "M" for millions, and "B" for billions US dollar.
If the signifier variable is empty, the number is damage in US dollar.
We can create new variables for property and crop damage (in US dollars) by multiplying the damage number with the appropriate power of ten:

```{r damage, echo=TRUE}
storms.df <- storms.df %>% 
    mutate_each(funs(revalue(., c("K" = "1e3", "M" = "1e6", "B" = "1e9"))), PROPDMGEXP, CROPDMGEXP) %>% 
    droplevels %>% 
    mutate(propdmg.multiplier = as.numeric(as.character(PROPDMGEXP)),
           cropdmg.multiplier = as.numeric(as.character(CROPDMGEXP))) %>% 
    mutate_each(funs(ifelse(is.na(.), 1, .)), propdmg.multiplier, cropdmg.multiplier) %>% 
    mutate(property.damage = PROPDMG * propdmg.multiplier,
           crop.damage = CROPDMG * cropdmg.multiplier) %>% 
    select(REFNUM:INJURIES, property.damage, crop.damage)
```

## Final Processed Data

The resulting `storms.df` data frame contains `r nrow(storms.df)` events and the following variables:

- `REFNUM`: Unique reference number for the event
- `year`: Year in which the event began
- `EVTYPE`: Event type
- `FATALITIES`: Number of fatalities due to the event
- `INJURIES`: Number of injuries due to the event
- `property.damage`: Property damage in US dollars due to the event
- `crop.damage`: Crop damage in US dollars due to the event

Table 1 shows the first 10 rows.

```{r storms.df, echo=TRUE}
kable(head(storms.df, n = 10), caption="Table 1. First 10 Rows of the Processed Dataset.")
```


# Results

This analysis addresses the following questions: Across the United States, which types of events

- are most harmful with respect to population health?
- have the greatest economic consequences?

## Population Health

Harm to population health in this analysis is measured in terms of numbers of fatalities and numbers of injuries due to severe weather events.
Beware that possible other types of harms  **cannot** be assessed based on this dataset.

Figure 1 shows total fatalities and injuries by event type.

```{r pop.health, echo=TRUE, fig.width=14, fig.height=8, fig.cap="Figure 1. Total fatalities and injuries by event type. A: Total numbers of fatalities/injuries for all event types, summed over all years between 1996 - 2011. Bars for the event types are sorted from greatest to smallest harm to population health (fatalities and injuries combined). B: Total numbers of fatalities/injuries by year for the 5 most harmful event types."}
harm.by.type <- storms.df %>% 
    group_by(EVTYPE) %>% 
    summarise(Fatalities = sum(FATALITIES), Injuries = sum(INJURIES)) %>% 
    arrange(desc(Fatalities), desc(Injuries))
harm.by.type$EVTYPE <- reorder(harm.by.type$EVTYPE, rowSums(harm.by.type[-1]))

harm.df <- harm.by.type %>% gather(Harm, Total, -EVTYPE)
p1 <- ggplot(harm.df, aes(x = EVTYPE, y = Total, fill = Harm))
p1 <- p1 + geom_bar(stat = "identity", position = 'stack') +
    xlab("Event Type") +
    ylab("Total number") + 
    ggtitle("A") +
    coord_flip() +
    theme_bw()

harm.by.type.year <- storms.df %>% 
    group_by(EVTYPE, year) %>% 
    summarise(Fatalities = sum(FATALITIES), Injuries = sum(INJURIES)) %>% 
    arrange(year, desc(Fatalities), desc(Injuries))
harm.by.type.year$EVTYPE <- reorder(harm.by.type.year$EVTYPE, rowSums(harm.by.type.year[3:4]))
top5 <- rev(levels(harm.by.type.year$EVTYPE))[1:5]

harm.df2 <- harm.by.type.year %>% 
    gather(Harm, Total, -c(EVTYPE, year)) %>% 
    filter(EVTYPE %in% top5)
p2 <- ggplot(harm.df2, aes(x = year, y = Total))
p2 <- p2 + geom_line(aes(colour = EVTYPE), size = 2, alpha = 0.6) +
    xlab("Year") +
    ylab("Total number") + 
    ggtitle("B") +
    facet_grid(Harm ~ ., scales = "free") +
    labs(colour = "Event Type") +
    theme_bw()

grid.arrange(p1, p2, ncol = 2)
```

Note that "UNMATCHED" events (cf. section [Cleaning Up Event Types](#cleaning-up-event-types)) account only for a minor fraction of overall fatalities/injuries.

Tornados, excessive heat, flood, thunderstorm wind, and lightning pose the greatest risk to population health. Tornados injure the most people by far, whereas excessive heat claims most fatalities.
In general, injury numbers are one order of magnitude larger than fatality numbers.
There is considerable variability across years, with differing patterns for fatalities and injuries.
The largest fluctuations with respect to fatalities occur with excessive heat and tornados, whereas for injuries, flood and tornado numbers vary the strongest.


## Economic Consequences

Figure  2 shows total damage (in Billion US dollars) to property and crop by event type.

```{r econ.conseq, echo=TRUE, fig.width=14, fig.height=8, fig.cap="Figure 2. Total damage (in Billion US dollars) to property and crop by event type. A: Total damage by event type, summed over all years between 1996 - 2011. Bars for the event types are sorted from highest to lowest damage (property and crop damage combined). B: Total damage by year for the 5 event types producing the most damage."}
damage.by.type <- storms.df %>% 
    group_by(EVTYPE) %>% 
    summarise(Property = sum(property.damage), 
              Crop = sum(crop.damage)) %>% 
    arrange(desc(Property), desc(Crop))
damage.by.type$EVTYPE <- reorder(damage.by.type$EVTYPE, rowSums(damage.by.type[-1]))

damage.df <- damage.by.type %>% gather(Damage, Total, -EVTYPE)
p3 <- ggplot(damage.df, aes(x = EVTYPE, y = Total/1e9, fill = Damage))
p3 <- p3 + geom_bar(stat = "identity", position = 'stack') +
    xlab("Event Type") +
    scale_y_continuous("Total Damage in Billion US Dollars") + 
    ggtitle("A") +
    coord_flip() +
    theme_bw()

damage.by.type.year <- storms.df %>% 
    group_by(EVTYPE, year) %>% 
    summarise(Property = sum(property.damage), 
              Crop = sum(crop.damage)) %>% 
    arrange(year, desc(Property), desc(Crop))
damage.by.type.year$EVTYPE <- reorder(damage.by.type.year$EVTYPE, rowSums(damage.by.type.year[3:4]))
top5 <- rev(levels(damage.by.type.year$EVTYPE))[1:5]

damage.df2 <- damage.by.type.year %>% 
    gather(Damage, Total, -c(EVTYPE, year)) %>% 
    filter(EVTYPE %in% top5)
p4 <- ggplot(damage.df2, aes(x = year, y = Total/1e9))
p4 <- p4 + geom_line(aes(colour = EVTYPE), size = 2, alpha = 0.6) +
    xlab("Year") +
    ylab("Total Damage in Billion US Dollars") + 
    ggtitle("B") +
    facet_grid(Damage ~ ., scales = "free") +
    labs(colour = "Event Type") +
    theme_bw()

grid.arrange(p3, p4, ncol = 2)
```

Note that "UNMATCHED" events (cf. section [Cleaning Up Event Types](#cleaning-up-event-types)) only account for a very minor fraction of the damage.

Economic consequences are mainly due to property damage, with flood, hurricane, storm surge, tornado, and hail events contributing the most over the years.
A notable exception are drought events which damage crops the most.
Variability across the years is mostly produced by flood and hurricane events.

